Category-based and location-based volitional covert attention affect memory at different timescales

### 2023-03-26 paper/main.tex
finished complete draft of discussion section

\textbf{JRM STOPPED HERE}

When participants held the focus of their category-based (face
versus place) and location-based (left versus right) attention for a sustained
interval, they judged stimuli they had seen as familiar when they overlapped
with respect to the features and locations they had attended. The increase in
familiarity was larger for attended-feature images than attended-location
images. The increase also extended to novel stimuli from the attended image
category. By contrast, when participants varied the focus of their
feature-based and location-based attention more rapidly, the boost in
familiarity for feature-matched stimuli was smaller than that for
location-matched stimuli, and did not extend to novel stimuli. Our findings
suggest that participants were able to more rapidly modulate their focus of
location-based attention than their focus of feature-based attention. The
tuning of location-based attention appears to be mediated by enhanced encoding
and faster processing at the attended location. The tuning of feature-based
attention appears to be mediated by a suppression in the encoding and
processing of unattended stimulus features. This suppression effect also
affects how new stimuli are processed, and it persists for a longer duration
following an interval when the focus of feature-based attention was held
constant over a longer duration. Taken together, our findings suggest that
feature-based and location-based attention are mediated by different mechanisms
and affect memory at different timescales (Fig.~\ref{fig:model}).

% \begin{figure*}[tp]
%   \centering \includegraphics[width=0.75\textwidth]{figures/discussion_schematic}

%   \caption{\textbf{Timecourse of encoding and retrieval effects of
%   feature-based and location-based attention.} The focus of location-based
%   attention (green) may be modulated quickly and rapidly enhances memory
%   encoding for nearby stimuli. The focus of feature-based attention (purple) is
%   modulated more slowly, and serves to suppress unattended stimulus features.
%   The effects of location-based attention on memory persist for longer than the
%   effects of feature-based attention. Sustained attention (darker shading)
%   yields more robust enhancement and suppression than shorter-term (variable)
%   attention (light shading). The vertical gray line on the left denotes the
%   duration of a single stimulus presentation (the upper bound by which
%   location-based attention begins to affect encoding, and the lower-bound by
%   which feature-based attention begins to affect encoding). The vertical gray
%   line on the right separates encoding from immediate subsequent retrieval of
%   the encoded information.}

% \label{fig:model}
% \end{figure*}

The notion that location-based attention operates at a faster timescale than
feature-based attention is supported by prior work on the deployment of visual
attention~\citep{SotoBlan04, StopEtal07}. Our findings that location-based
attention enhances the processing of attended stimuli whereas feature-based
attention suppresses the processing of unattended stimuli is also consistent
with prior work on location-based attention~\citep[e.g.,][]{IttiKoch01} and
feature-based attention~\citep[e.g.,][]{MoheEtal14}. Our finding that people
better remember attended stimuli also follows prior work on interactions
between attention and memory~\citep{PallWagn02, ChunTurk07, AlyTurk16,
AlyTurk17, WittEtal18, MorrEtal14, BaleEtal21}. Whereas much of this prior work
focused on elucidating the neural basis of these interactions, our work extends
these prior studies by elucidating the specific and separable behavioral
impacts of feature-based attention (inhibition with a slow onset) and
location-based attention (enhancement with a fast onset) on subsequent memory.
Both of these effects persisted throughout the 2~min memory phases of both
experiments. Therefore future work is needed to elucidate the longevity of
these effects beyond 2 minutes.

Another important area for future study concerns how the flow of information
between different brain structures is modulated according to the focus of
volitional attention-- particularly with respect to pathways from primary
sensory regions (e.g., V1, A1) to regions implicated in encoding ongoing
experiences into memory (e.g, medial temporal lobe structures such as the
hippocampus and entorhinal cortex, prefrontal cortex, etc.). For example,
several studies suggest that attention serves to modulate the \textit{gain} of
specific neural circuits~\citep{TreuTruj99, ChanEtal02, EldaEtal13,
SaliThie00}, effectively facilitating or inhibiting the flow of specific neural
representations~\citep{VartEtal07, LaRoEtal14}. Prior work suggests that
feature-based attention may be supported by changes in connectivity with the
thalamus~\citep{Schn11}, whereas location-based attention may be supported by
changes in connectivity with primary visual cortex~\citep{NoudEtal10}. That
feature-based and location-based attention are mediated by different brain
structures may explain why these different aspects of attention operate on
different timescales and affect memory differently. A strong test of this
hypothesis would entail directly measuring neural activity patterns as people
modulate their focus of attention (e.g., using functional magnetic resonance
imaging or electroencephalography), and then using neural decoding
approaches~\citep[e.g.,][]{HaxbEtal01, NormEtal06b, MannEtal18} to follow how
neural representations of attended (or unattended) stimuli are transfered from
primary sensory regions, to higher order sensory regions, to memory areas. If
the effects of attention on memory are mediated by changes in network dynamics,
the transmission rates of the representations of attended stimuli from primary
sensory regions to memory areas should be facilitated relative to the
transmission rates of unattended stimuli. Further, variability in these neural
changes (e.g., as a participant focuses their attention with more or less
success) should track with behavioral measures of memorability.

Which aspects of our ongoing experiences we choose to attend affects how we
process and remember those experiences later. Different forms of
attention---e.g., to specific features or spatial locations---operate and
affect memory at different timescales, and are likely mediated by different
brain networks. Elucidating the behavioral and neural consequences of
volitional changes in attention is central to discovering how our thoughts,
feelings, goals, and situational understanding fluctuate from moment to moment.

>>>>

The notion that location-based attention operates at a faster timescale than
category-based attention is supported by prior work on the deployment of visual
attention~\citep{SotoBlan04, StopEtal07}. Our findings that location-based
attention enhances the processing of attended stimuli whereas category-based
attention suppresses the processing of unattended stimuli is also consistent
with prior work on location-based attention~\citep[e.g.,][]{IttiKoch01} and
category-based attention~\citep[e.g.,][]{MoheEtal14}. Our finding that people
better remember attended stimuli also follows prior work on interactions
between attention and memory~\citep{PallWagn02, ChunTurk07, AlyTurk16,
AlyTurk17, WittEtal18, MorrEtal14, BaleEtal21}. Whereas much of this prior work
focused on elucidating the neural basis of these interactions, our work extends
these prior studies by elucidating the specific and separable behavioral
impacts of location-based attention (enhancement with a fast onset) and
category-based attention (inhibition with a slow onset) and on subsequent
recognition memory. Both of these effects persisted throughout the 2~min memory
phases of both experiments. Therefore future work is needed to elucidate the
longevity of these effects beyond 2 minutes.

Another important area for future study concerns how the flow of information
between different brain structures is modulated according to the focus of
volitional attention-- particularly with respect to pathways from primary
sensory regions (e.g., V1, A1) to regions implicated in encoding ongoing
experiences into memory (e.g, medial temporal lobe structures such as the
hippocampus and entorhinal cortex, prefrontal cortex, etc.). For example,
several studies suggest that attention serves to modulate the \textit{gain} of
specific neural circuits~\citep{TreuTruj99, ChanEtal02, EldaEtal13,
SaliThie00}, effectively facilitating or inhibiting the flow of specific neural
representations~\citep{VartEtal07, LaRoEtal14}. Prior work suggests that
category-based attention may be supported by changes in connectivity with the
thalamus~\citep{Schn11}, whereas location-based attention may be supported by
changes in connectivity with primary visual cortex~\citep{NoudEtal10}. That
category-based and location-based attention are mediated by different brain
structures may explain why these different aspects of attention operate on
different timescales and affect memory differently. A strong test of this
hypothesis would entail directly measuring neural activity patterns as people
modulate their focus of attention (e.g., using functional magnetic resonance
imaging or electroencephalography), and then using neural decoding
approaches~\citep[e.g.,][]{HaxbEtal01, NormEtal06b, MannEtal18, OwenEtal21} to
follow how neural representations of attended (or unattended) stimuli are
transfered from primary sensory regions, to higher order sensory regions, to
memory areas. If the effects of attention on memory are mediated by changes in
network dynamics, the transmission rates of the representations of attended
stimuli from primary sensory regions to memory areas should be facilitated
relative to the transmission rates of unattended stimuli. Further, variability
in these neural changes (e.g., as a participant focuses their attention with
more or less success) should track with behavioral measures of memorability.

Which aspects of our ongoing experiences we choose to attend affects how we
process and remember those experiences later. Different forms of
attention---e.g., to specific stimulus categories or spatial
locations---operate and affect memory at different timescales, and are likely
mediated by different brain networks. Elucidating the behavioral and neural
consequences of volitional changes in attention is central to discovering how
our thoughts, feelings, goals, and situational understanding fluctuate from
moment to moment.
